Transcriptomic Analysis of Expressions of Biomarkers and Their Prognostic Value in Head and Neck Squamous Cell Carcinoma

### 2021-06-27 Cover_letter_JPM.tex
Updates from Overleaf

% [2021/06/13-17] GSE65858
%CAMK2N1, IL19, and FCGBP is validated by
%candidate_fdr100
%PvalueTex consensus with GSE2837 to yield 3 candidates
In the study of GSE65858 cohort with median cutoffs, CAMK2N1, IL19, and FCGBP (3 of the 20 genes), keep ahead of the curve by their FDR-corrected \textit{P} value ($< 0.05$), and Cox'HR ($>1.8$ or $<0.65$) (please see Table \ref{tab:table_top3_GSE65858} listed below,
% (ok) (FDR KM \textit{P} value) <= # only MSMB KM FDR-adjusted P value = 0.002 
% Consensus 3 genes in Kaplan-Meier survival and Cox's model (dataset: GSE65858)
%Gene.Symbol KM.Pvalue   FDR.KMpvalue    uni_HR  FDR.Pvalue(in TCGA)
%CAMK2N1 0.00687 0.03782500  1.814   1.628308e-05
%FCGBP   0.01020 0.03886709  0.573   4.827833e-05
%IL19    0.03390   0.04738901    0.630   6.543871e-06
%IL19    0.00269  0.03068073 0.630  6.543871e-06
%candidate_fdr100
and also see Figure \ref{fig:hazards534}).
However, the significance of the other 17 genes is replaced by genes such as DUSP6, MSMB, and RBM11 (please see Figure \ref{fig:hazards534}).
This GSE65858 cohort has been applied for selection to generate our candidate genes: \textcolor{red}{CAMK2N1}, \textcolor{red}{IL19} (two probes), and , \textcolor{red}{FCGBP}.
Moreover, there are 22 genes, which have greater hazard ratio ($> 1.8$ or $< 0.6$) in GSE65858 cohort (Figure \ref{fig:hazards534}), drop their hazard ratio between 0.6 and 1.5 in the study of TCGA HNSCC (Figure \ref{fig:hazards3}).
Thus, there is a consensus between the TCGA and GSE65858 cohorts that CAMK2N1, IL19, and FCGBP are significant candidates for the HNSCC biomarker.
%The consensus between the TCGA and GSE65858 cohorts in Kaplan-Meier survival and Cox's model

@@

%When we increase the sample size, the standard error decrease
%, or increase the difference between the sample statistic and hypothesized parameter, the p value decreases, thus making it more likely that we reject the null hypothesis.
The larger the sample the more likely a difference to be accepted (i.e. \textit{P} value $< 0.05$), due to the decrease of the standard error\cite{Sullivan2012}.
However, the small effect size (non-zero) is often meaningless (substantive insignificance).
%With a sufficiently large sample, a statistical test will always reveal a significant difference, unless the effect size is exactly zero.
%, that is, when the effect size is exactly zero; 
%yet very small differences, even if significant, are often meaningless. 
%Thus, reporting only the significant P value for an analysis is not adequate for readers to fully understand the results.
%If the magnitude of effect is small and clinically unimportant, the p-value can be "significant" if the sample size is large.
Conversely, an effect size can be large, but fail to get statistical significance if the sample size is small. In this scenario, the \textit{P} value tells the effect does not exist.

The \textit{P} value could be also impacted by error.
\begin{outline}
\1 A random error, defined as variability in data, is not considered a bias but rather occurs random across the entire study population and can distort the measurement process (e.x. RNA-Seq experiments).
The larger sample size could reduce the random error. 
\1 A systematic error is a bias, which includes selection bias, information bias, and confounding.
It could deleteriously impact the statistical significance.
The larger sample size could not affect the systematic error.
\end{outline}
Thus, if there is no error in two study groups, and the sample size is the same, the group, which has a larger effect size, will has small \textit{P} value\cite{Thiese2016}.

If a skewed cutoff that split, for example, 425 vs 75 between the two groups, it will also get statistical significance by increasing effect size artificially.
% https://stats.stackexchange.com/questions/162359/how-can-you-have-p-0-0001-with-a-sample-size-of-89-and-a-control-group-of-52
%As you can see, even with only 12 events (fustat = 1) that split 10 vs 2 between the two dummy groups you get a log-rank p-value < 0.0001.
%effect size, sample size, cutoff 影響 p value!

@@

2) \\
( ) % pvalueTex is essential
There is benefit of using sliding-window cutoff method in survival analysis of TCGA HNSCC cohort.
We have compared the results of cutoff at optimal by sliding-window or just at median of gene expression.
It shows
the gene number with unadjusted \textit{P} value $<0.05$, which is 6284 vs. 3118, respectively.
After FDR-correction, it shows 967 vs. 209, respectively.
% HNSCC_OS_marginS_THREE_pvalue005_noCancerGene
%Result by optimal cutoff: total 9416 plots,
%6284 are raw KM P value < 0.05
%534?? genes are FDR KM P value <0.05
%967 genes which HR is greater than 1.5 or less than 0.5.
%Result by median cutoff: total 20080 genes, 
%3118 are raw KM P value < 0.05; 
%209 genes are FDR KM P value <0.05
The sliding-window cutoff method could catch more potential candidates (\textit{P} values far less than $0.001$) for Cox's modeling at next stage in our workflow.
These preliminary candidates must be validated by Bonferroni correction and the other independent cohorts (GSE65858, GSE2837) to prevent the false discovery.

@@

We will explain above mentioned by two examples.
%***NDFIP1  % 給他們一個機會
Reviewing the special cases of gene, such as NDFIP1, DKC1, PNMA5, and NPB,
we have noticed that NDFIP1, with a \textit{P} value of 0.05 at 50\% quantile cutoff, could achieve a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff.
NDFIP1 has the FDR-corrected \textit{P} value as \num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.0001074233} (please see supplementary figure S1). % it is \ref{fig:pvaluePlot_NDFIP1
%Moreover, the gene NDFIP1, one of our 20 preliminary candidates, has a \textit{P} value (around $0.05$) at 50\% quantile cutoff, achieving a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff (please see Figure \ref{fig:pvaluePlot_NDFIP1}, and also in supplementary figure S1).
%After the FDR correction, NDFIP1 still has \textit{P} value of \num[round-precision=3, round-mode=figures, scientific-notation=true]{0.0001074233}.
% the effect size
Although it has failed to compete as a candidate by less effect size (HR = 1.33) in GSE65858 cohort, compared with one (HR = 2.027) in TCGA cohort%.
, and the other (HR = 1.42) in GSE2837 cohort.
% (ok) Answer 2-1 加入 discussion
%According to the results of sliding-window cutoff finding, the preliminary biomarkers have FDR-corrected \textit{P} values from \num{1e-4} to 0.04 which have been grouped by cutoffs between 50\% and 70\% quantile (Please see Figure \ref{fig:figure4}; it is also shown in manuscript as updated "Figure 4": top 3 candidate biomarkers, CAMK2N1, IL19, and FCGBP).

@@

% ****** IL19 必須要靠 pvalueTex 才能發現
%*** IL19 is excluded by this way :-) (special S-curve P value plot)
The other example is one of our candidate: IL19.
It has a \textit{P} value plot with acute S-curve bending at the median zone, that lets the FDR-KM \textit{P} value has large difference between median cutoff (\num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.1154315}) and optimal cutoff (\num{6.54e-6}).
The sliding-window selection of cutoffs could boost their statistic significance to pass the correction by both FDR and Bonferroni methods.
%> gene_labeled_golden[!gene_labeled_golden %in% fdr100_KMsurvival$Gene.Symbol[fdr100_KMsurvival$FDR.Pvalue<0.05]]
%[1] "IL19"
Fortunately, IL19 has passed to become a candidate by better effect size (HR = 0.472 in TCGA cohort), and also by the validation with the other independent cohort (HR = 0.630 in GSE65858 cohort%).
, and HR = 0.84 in GSE2837).

>>>>

% [2021/06/13-17] GSE65858
%CAMK2N1, IL19, and FCGBP is validated by
%candidate_fdr100
%PvalueTex consensus with GSE2837 to yield 3 candidates
In the study of GSE65858 cohort with median cutoffs, CAMK2N1, IL19, and FCGBP (3 of the 20 genes), keep ahead of the curve by their FDR-corrected \textit{P} value ($< 0.05$), and Cox'HR ($>1.8$ or $<0.65$) (please see Table \ref{tab:table_top3_GSE65858} listed below,
% (ok) (FDR KM \textit{P} value) <= # only MSMB KM FDR-adjusted P value = 0.002 
% Consensus 3 genes in Kaplan-Meier survival and Cox's model (dataset: GSE65858)
%Gene.Symbol KM.Pvalue   FDR.KMpvalue    uni_HR  FDR.Pvalue(in TCGA)
%CAMK2N1 0.00687 0.03782500  1.814   1.628308e-05
%FCGBP   0.01020 0.03886709  0.573   4.827833e-05
%IL19    0.03390   0.04738901    0.630   6.543871e-06
%IL19    0.00269  0.03068073 0.630  6.543871e-06
%candidate_fdr100
and also see Figure \ref{fig:hazards534}).
However, the significance of the other 17 genes is replaced by genes such as DUSP6, MSMB, and RBM11 (please see Figure \ref{fig:hazards534}).
%This GSE65858 cohort has been applied for selection to generate our candidate genes: \textcolor{red}{CAMK2N1}, \textcolor{red}{IL19} (two probes), and , \textcolor{red}{FCGBP}.
Moreover, there are 22 genes, which have greater hazard ratio ($> 1.8$ or $< 0.6$) in GSE65858 cohort (Figure \ref{fig:hazards534}), drop their hazard ratio between 0.6 and 1.5 in the study of TCGA HNSCC (Figure \ref{fig:hazards3}).
Thus, there is a consensus between the TCGA and GSE65858 cohorts that CAMK2N1, IL19 (two probes), and FCGBP are significant candidates for the HNSCC biomarker.
%The consensus between the TCGA and GSE65858 cohorts in Kaplan-Meier survival and Cox's model

@@

%When we increase the sample size, the standard error decrease
%, or increase the difference between the sample statistic and hypothesized parameter, the p value decreases, thus making it more likely that we reject the null hypothesis.
The larger the sample the more likely a difference to be noticed (i.e. \textit{P} value $< 0.05$), due to the decrease of the standard error\cite{Sullivan2012}.
However, the small effect size (non-zero) is often meaningless (substantive insignificance).
%With a sufficiently large sample, a statistical test will always reveal a significant difference, unless the effect size is exactly zero.
%, that is, when the effect size is exactly zero; 
%yet very small differences, even if significant, are often meaningless. 
%Thus, reporting only the significant P value for an analysis is not adequate for readers to fully understand the results.
%If the magnitude of effect is small and clinically unimportant, the p-value can be "significant" if the sample size is large.
Conversely, an effect size can be large, but fail to get statistical significance if the sample size is small. 
%In this scenario, the \textit{P} value tells the effect does not exist.

The \textit{P} value could be also impacted by error, which includes:
\begin{outline}
\1 A random error, defined as variability in data, is not considered a bias but rather occurs random across the entire study population and can distort the measurement process (e.x. RNA-Seq experiments).
The larger sample size could reduce the random error. 
\1 A systematic error is a bias, which includes selection bias, information bias, and confounding.
It could deleteriously impact the statistical significance.
The larger sample size could not affect the systematic error.
\end{outline}
% effect size is large (whatever is true or artificial), the p value is small
Thus, if there is no error in two study groups, and the sample size is the same, the group, which has a larger effect size, will has small \textit{P} value\cite{Thiese2016}.
If a skewed cutoff that split, for example, 425 vs 75 between the two groups, it will also get statistical significance by increasing effect size artificially.
% https://stats.stackexchange.com/questions/162359/how-can-you-have-p-0-0001-with-a-sample-size-of-89-and-a-control-group-of-52
%As you can see, even with only 12 events (fustat = 1) that split 10 vs 2 between the two dummy groups you get a log-rank p-value < 0.0001.
%effect size, sample size, cutoff 影響 p value!

@@

%2) \\
%(ok) % pvalueTex is essential
There is a benefit for using sliding-window cutoff method (avoiding a skewed cutoff) in Kaplan-Meier analysis of TCGA HNSCC cohort at the beginning.
We have compared the results of cutoff at optimal \textit{P} value by sliding-window or just at median of gene expression.
It shows that the number of genes with unadjusted \textit{P} value $<0.05$, which is 6284 vs. 3118, respectively. After FDR-correction, it has reduced as 967 vs. 209, respectively.
% HNSCC_OS_marginS_THREE_pvalue005_noCancerGene
%Result by optimal cutoff: total 9416 plots,
%6284 are raw KM P value < 0.05
%534?? genes are FDR KM P value <0.05
%967 genes which HR is greater than 1.5 or less than 0.5.
%Result by median cutoff: total 20080 genes, 
%3118 are raw KM P value < 0.05; 
%209 genes are FDR KM P value <0.05
The sliding-window cutoff method could catch more potential candidates, which have \textit{P} values far less than $0.001$, for subsequent Cox's modeling.
That is because of proper selected cutoff improving the statistical significance.
Then, these preliminary candidates will have an opportunity to be carefully selected by Bonferroni correction, their effect size (HR), and the other independent cohorts (GSE65858, GSE2837) to prevent the false discovery.

@@

We will explain above mentioned by two examples.
%***NDFIP1  % 給他們一個機會
Reviewing the special cases of gene, such as NDFIP1, DKC1, PNMA5, and NPB,
we have noticed that NDFIP1, with a \textit{P} value of 0.05 at 50\% quantile (median) cutoff, could achieve a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff.
NDFIP1 has the FDR-corrected \textit{P} value as \num[round-precision=3, round-mode=figures,
scientific-notation=true]{0.0001074233} (please see supplementary figure S1, a "W"-shaped \textit{P} value plot). % it is \ref{fig:pvaluePlot_NDFIP1
%Moreover, the gene NDFIP1, one of our 20 preliminary candidates, has a \textit{P} value (around $0.05$) at 50\% quantile cutoff, achieving a \textit{P} value of \num{2.62e-6} at 70\% quantile cutoff (please see Figure \ref{fig:pvaluePlot_NDFIP1}, and also in supplementary figure S1).
%After the FDR correction, NDFIP1 still has \textit{P} value of \num[round-precision=3, round-mode=figures, scientific-notation=true]{0.0001074233}.
% the effect size
Then it has been excluded as a candidate by less effect size (HR = 1.33) in GSE65858 cohort, compared with one (HR = 2.027) in TCGA cohort%.
, and the other (HR = 1.42) in GSE2837 cohort.
% (ok) Answer 2-1 加入 discussion
%According to the results of sliding-window cutoff finding, the preliminary biomarkers have FDR-corrected \textit{P} values from \num{1e-4} to 0.04 which have been grouped by cutoffs between 50\% and 70\% quantile (Please see Figure \ref{fig:figure4}; it is also shown in manuscript as updated "Figure 4": top 3 candidate biomarkers, CAMK2N1, IL19, and FCGBP).

@@

% ****** IL19 必須要靠 pvalueTex 才能發現
%*** IL19 is excluded by this way :-) (special S-curve P value plot)
The other example is one of our candidate: IL19.
It has a \textit{P} value plot with acute S-curve bending at the median zone, that lets the FDR-adjusted \textit{P} value has large difference between 50\% quantile cutoff (FDR 0.115)
%(\num[round-precision=3, round-mode=figures, scientific-notation=true]{0.1154315}) 
and optimal 48\% quantile cutoff (FDR \num{6.54e-6}).
This optimal cutoff method could boost it's statistic significance to pass the correction by both FDR and Bonferroni methods.
%> gene_labeled_golden[!gene_labeled_golden %in% fdr100_KMsurvival$Gene.Symbol[fdr100_KMsurvival$FDR.Pvalue<0.05]]
%[1] "IL19"
Fortunately, IL19 has competed to become a candidate by its better effect size (HR = 0.472 in TCGA cohort), and also by the validation with the other independent cohort (HR = 0.630 in GSE65858 cohort).
%, and HR = 0.84 in GSE2837).
